---
title: "Prediction Model - Exercise Manner"
author: "Brandon Hu"
date: "March 20, 2016"
output: html_document
---

###Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 

This project goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

More information is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (see the section on the Weight Lifting Exercise Dataset).

###Data Preparation
***
Download and load test and training data.

```{r}
# Data Loading
#-------------
# trainingDataUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# testDataUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# Loading the data
# download.file(trainingDataUrl,destfile = "pml_train.csv")
# download.file(testDataUrl,destfile = "pml_test.csv")

testingData = read.csv("pml-training.csv", na.strings=c("", "NA", "NULL"))
dim(testingData)

trainingData = read.csv("pml-training.csv", na.strings=c("", "NA", "NULL"))
dim(trainingData)
```

####Pre-screening Data
There are several approaches for reducing the number of predictors.

-Remove variables that we believe have too many NA values.
```{r}
training.dena <- trainingData[ , colSums(is.na(trainingData)) == 0]
dim(training.dena)
```
-Remove unrelevant variables There are some unrelevant variables that can be removed as they are unlikely to be related to dependent variable.

```{r}
remove = c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'new_window', 'num_window')
training.dere <- training.dena[, -which(names(training.dena) %in% remove)]
dim(training.dere)
```
-Check the variables that have extremely low variance (this method is useful nearZeroVar() )

```{r}
#install caret using: install.packages('caret', repos='http://cran.rstudio.com/')
#suppress messages and warnings if any from loading 'caret' package
suppressWarnings(suppressMessages(library(caret)))

# only numeric variabls can be evaluated in this way.
zeroVar= nearZeroVar(training.dere[sapply(training.dere, is.numeric)], saveMetrics = TRUE)
training.nonzerovar = training.dere[,zeroVar[, 'nzv']==0]
dim(training.nonzerovar)
```
-Remove highly correlated variables 90% (using for example findCorrelation() )
```{r, echo=TRUE}
# only numeric variabls can be evaluated in this way.
corrMatrix <- cor(na.omit(training.nonzerovar[sapply(training.nonzerovar, is.numeric)]))
dim(corrMatrix)
# there are 52 variables.
corrDF <- expand.grid(row = 1:52, col = 1:52)
corrDF$correlation <- as.vector(corrMatrix)
levelplot(correlation ~ row+ col, corrDF)
```
We are going to remove those variable which have high correlation.

```{r, echo=TRUE}
removecor = findCorrelation(corrMatrix, cutoff = .90, verbose = FALSE)
training.decor = training.nonzerovar[,-removecor]
dim(training.decor)
```
We get 19622 samples and 46 variables.

####Split data to training and testing for cross validation
```{r}
inTrain <- createDataPartition(y=training.decor$classe, p=0.7, list=FALSE)
training <- training.decor[inTrain,]; testing <- training.decor[-inTrain,]
dim(training);
dim(testing)
```
###Analysis
***
###Regression Tree

Now we fit a tree to these data, and summarize and plot it. First, we use the 'tree' package. It is much faster than 'caret' package.

```{r}
suppressWarnings(suppressMessages(library(tree)))
set.seed(12345)
tree.training=tree(classe~.,data=training)
summary(tree.training)
plot(tree.training)
text(tree.training,pretty=0, cex =.8)
```
This is a bushy tree, and we are going to prune it.

###Rpart Model
```{r}
#To prevent e1071 package error, install.packages('e1071', dependencies=TRUE)
modFit <- train(classe ~ .,method="rpart",data=training)
print(modFit$finalModel)
```
To generate prettier plots RPart decision tree plotter,
```{r}
#install.packages('rattle') from R and RTK2 from R
suppressWarnings(suppressMessages(library(rattle)))
fancyRpartPlot(modFit$finalModel)
```
The result from 'caret' 'rpart' package is close to 'tree' package. 

###Cross Validation
We are going to check the performance of the tree on the testing data by cross validation.
```{r}
tree.pred=predict(tree.training,testing,type="class")
predMatrix = with(testing,table(tree.pred,classe))
sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate
```

The 0.70 is not very accurate.
```{r}
tree.pred=predict(modFit,testing)
predMatrix = with(testing,table(tree.pred,classe))
sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate
```
The 0.50 from 'caret' package is much lower than the result from 'tree' package.

###Pruning tree

This tree was grown to full depth, and might be too variable. We now use Cross Validation to prune it.
```{r}
cv.training=cv.tree(tree.training,FUN=prune.misclass)
cv.training
plot(cv.training)
```
It shows that when the size of the tree goes down, the deviance goes up. It means the 21 is a good size (i.e. number of terminal nodes) for this tree. We do not need to prune it.

Suppose we prune it at size of nodes at 18.
```{r}
prune.training=prune.misclass(tree.training,best=18)
```
Now lets evaluate this pruned tree on the test data.

```{r}
tree.pred=predict(prune.training,testing,type="class")
predMatrix = with(testing,table(tree.pred,classe))
sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate
```
0.66 is a little less than 0.70, so pruning did not hurt us with repect to misclassification errors, and gave us a simpler tree. We use less predictors to get almost the same result. By pruning, we got a shallower tree, which is easier to interpret.

The single tree is not good enough, so we are going to use bootstrap to improve the accuracy. We are going to try random forests.

###Random Forests
Random forests build lots of bushy trees, and then average them to reduce the variance.
```{r}
require(randomForest)
set.seed(12345)
```
Lets fit a random forest and see how well it performs.
```{r}
rf.training=randomForest(classe~.,data=training,ntree=100, importance=TRUE)
rf.training
varImpPlot(rf.training)
```
we can see which variables have higher impact on the prediction.

###Out-of Sample Accuracy
***
Our Random Forest model shows OOB estimate of error rate: 0.72% for the training data. Now we will predict it for out-of sample accuracy.

Now lets evaluate this tree on the test data.

```{r}
tree.pred=predict(rf.training,testing,type="class")
predMatrix = with(testing,table(tree.pred,classe))
sum(diag(predMatrix))/sum(as.vector(predMatrix)) # error rate
```
0.99 means we got a very accurate estimate.

No. of variables tried at each split: 6. It means every time we only randomly use 6 predictors to grow the tree. Since p = 43, we can have it from 1 to 43, but it seems 6 is enough to get the good result

###Conclusion
Now we can predict the testing data from the website.
```{r}
answers <- predict(rf.training, testingData)
answers
```
Those answers are going to submit to website for grading. It shows that this random forest model did a good job.